**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**

### 2021-03-01 Manuscripts/Manuscript_AM_v3.Rmd
Updated manuscript

In this model $y_{ijt}$ is the response by subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively. The _random effect_ $\mu_{ij}$, accounts for variability in each subject within each group by allowing different values for each subject. Independent errors $\varepsilon_{tij}$ represent random variation not explained by the _fixed_ or _random_ effects. 
In a  biomedical research context, suppose two treatments groups are used in a study (e.g., "placebo" vs. "novel drug" or "saline" vs. "chemotherapy"). Then, the group terms in Equation \@ref(eq:linear-model) can be written as below with $j=0$ representing the first treatment group (Group A) and $j=1$ representing the second treatment group (Group B). The linear models then can be expressed as

\begin{equation}
y_{ijt} = \begin{cases}
\beta_0 + \beta_1\times time_{t}+\mu_{ijt}+\varepsilon_{ijt}   & \mbox{if Group A}\\
\beta_0 + \beta_1 \times time_{t} + \beta_2 \times treatment_{j} +\beta_3 \times time_{t} \times treatment_{j}+\mu_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
\end{cases}
(\#eq:ANOVA-by-group)
\end{equation}

To further simplify the expression, substitute $\widetilde{\mu}=\beta_0+\beta_{2}$ and $\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}$ in the equation for Group B. This substitution allows for a different intercept and slope for Group B. The model is written as

@@

Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different intercepts and slopes per group). In other words, When it is stated that an rm-ANOVA model "expects" a linear relationship between the covariates and the response, this means that either presented as Equation \@ref(eq:linear-model), Equation \@ref(eq:ANOVA-by-group) or Equation \@ref(eq:ANOVA-lines), an rm-ANOVA model is only able to accommodate linear patterns in the data. If the data show non-linear behavior, the rm-ANOVa model will approximate this behavior with non-parallel lines. Because the construction of the model is essentially the same in LMEMs, the linearity assumption holds for these models, although it is possible to fit models with different slopes at the subject-level[@pinheiro2006].

When the data does not follow a linear trend, the fit that an rm-ANOVA or LMEM produces may not accurately represent the trends in the data. To demonstrate this inaccurate representation, longitudinal data where a normally distributed response of two groups of 10 subjects each is simulated, and a rm-ANOVA model  (Equation \@ref(eq:linear-model)) is fitted to the data using $\textsf{R}$[@r] and the package _nlme_[@nlme] (code in the Appendix). Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes (negative slope for Group 1 and positive slope for Group 2 ). In the second case, a quadratic function is used for the mean response per group (concave down for Group 1 and concave up for Group 2 ). From the mean response per group (linear or quadratic), individual responses for each subject within each group are simulated using covariance matrices for with compound symmetry (constant variance across time) or independent errors (errors that are generated from a a normal distribution but are not constant over time). Thus, the response per subject corresponds to the mean response per group plus the error (compound symmetry or independent). To this data, an rm-ANOVA model identical as Equation  \@ref(eq:linear-model) is fitted. The resulting simulated data, simulated errors, individual response and fitted parameters from the model are plotted.

@@

In Figure \@ref(fig:linear-models)  panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response. In panel G,because the errors are independent, there is no expectation that responses are likely to follow a trend.  It is clear from \@ref(fig:linear-models) fit produced by the rm-ANOVA model is good as the predictions for the mean response are reasonably close to the "truth" of the simulated data.In particular,

>>>>

In this model $y_{ijt}$ is the response for subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively. The _random effect_ $\mu_{ij}$ is assumed to be independently normally distributed with mean zero and variance $\sigma^2_\mu$, which can be expressed as $\mu_{ij} \sim N(0,\sigma^2_\mu)$. This _random effect_ accounts for variability in each subject within each group by allowing different values for each subject. Independent errors $\varepsilon_{tij}$ represent random variation not explained by the _fixed_ or _random_ effects, and are assumed to be $\sim N(0,\sigma^2)$ . 
In a  biomedical research context, suppose two treatments groups are used in a study (e.g., "placebo" vs. "novel drug" or "saline" vs. "chemotherapy"). Then, the group terms in Equation \@ref(eq:linear-model) can be written as below with $treatment_j=0$ representing the first treatment group (Group A) and $treatment_j=1$ representing the second treatment group (Group B). The linear models then can be expressed as

\begin{equation}
y_{ijt} = \begin{cases}
\beta_0 + \beta_1\times time_{t}+\mu_{ijt}+\varepsilon_{ijt}   & \mbox{if Group A}\\
\beta_0 + \beta_1 \times time_{t} + \beta_2 \times +\beta_3 \times time_{t} \times +\mu_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
\end{cases}
(\#eq:ANOVA-by-group)
\end{equation}

To further simplify the expression, substitute $\widetilde{\mu}=\beta_0+\beta_{2}$ and $\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}$ in the equation for Group B. This substitution allows for a different intercept and slope for Groups A and B. The model is then written as

@@

Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different intercepts and slopes per group). In other words, the rm-ANOVA model "expects" a linear relationship between the covariates and the response, this means that either presented as Equation \@ref(eq:linear-model), Equation \@ref(eq:ANOVA-by-group) or Equation \@ref(eq:ANOVA-lines), an rm-ANOVA model is only able to accommodate linear patterns in the data. If the data show non-linear behavior, the rm-ANOVA model will approximate this behavior with non-parallel lines. Because the construction of the model is essentially the same in LMEMs, the linearity assumption holds for these models, although it is possible to fit models with different slopes at the subject-level[@pinheiro2006].

#Simulation experiment on linearity and covariance structures

To demonstrate the limitations of rm-ANOVA for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each. To each group, an rm-ANOVA model (Equation \@ref(eq:linear-model)) is fitted using $\textsf{R}$[@r] and the package _nlme_[@nlme] (code in the Appendix). 
Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group (Figure \@ref(fig:linear-models, A)) .In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \@ref(fig:quadratic-response),A). From the mean response per group (linear or quadratic), individual responses for each subject within each group are simulated using covariance matrices with compound symmetry (constant variance across time) or independent errors (errors that are generated from a a normal distribution that are not constant over time). Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (either compound symmetry or independent, which appear in panels A and E, respectively in both Figure \@ref(fig:linear-models) and Figure \@ref(fig:quadratic-response)) . To this data, an rm-ANOVA model identical as Equation  \@ref(eq:linear-model) is fitted. The resulting simulated data, simulated errors, individual response and fitted parameters from the model are plotted.

@@

In Figure \@ref(fig:linear-models)  panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value of the random error $\varepsilon $ either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response. In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.  It is clear from Figure \@ref(fig:linear-models) that the fit produced by the rm-ANOVA model is good as the predictions for the mean response are reasonably close to the "truth" of the simulated data.
